### 2008-12-03 htdocs/sections.tex
Added initial notes on surfactant free energy
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@444 40876441-3496-0410-a613-a2fca35b2ae2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.6 2008-11-10 11:23:58 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The following apply for the symmetric binary fluid problem with
order parameter $\phi$ and free energy (see section X) \cite{viv}
\begin{equation}
 F[\phi] = 
\int dr \left(
{\scriptstyle \frac{1}{2}}A\phi^2
+ {\scriptstyle \frac{1}{4}}B\phi^4
+ {\scriptstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right)
\end{equation}

@@

The parameter $\kappa$

\inputkey{C}

The surface free energy parameter {C}. Currently unused.

\inputkey{H}

The surface free energy parameter {H}. Currently unused.

@@

Sets the order parameter mobility in the LB framework.

@@

The mean compositional order parameter roughly $-0.5 < \phi < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

@@

An analtyical expression is also available for the conductance
of a square or rectugular capillary. For an infinite capillary
of square cross section width $w \times h$ [WHICH IS LONGER?],
the volume flux per unit area $J$ is expected to be
\cite{papanastasiou,edo1}
\begin{equation}
J = - \frac{1}{3\eta} \frac{\partial p}{\partial x} (h/2)^2
\left[
1 - 6(h/w) \sum_{k=1}^{\infty} \frac{\tanh(\alpha_k w/h)}{\alpha_k^5} 
\right]
\end{equation}
where $\alpha_k = (2k - 1)\pi/2$. The pressure gradient
$-\partial p / \partial x$ is replaced by the body force $\rho g$ and
one can difine the viscosity-independnet conductance $C$ via
$J = C\rho g / \eta$. The calculation procedes as above.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  sections.tex
%
%  Content for Ludwig documentation.
%
%  $Id: sections.tex,v 1.7 2008-12-03 20:19:27 kevin Exp $
%
%  Edinburgh Soft Matter and Statistical Physics Group and
%  Edinburgh Parallel Computing Centre
%
%  Kevin Stratford (kevin@epcc.ed.ac.uk)
%  (c) 2008 The University of Edinburgh
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The are currently three choices of free energy relating to
various systems.

\subsubsection{Symmetric binary fluid}

\inputkey{free\_energy symmetric}

The following applies for the binary fluid problem with
compositional order parameter $\phi$ and free energy
(excluding the term in the density $\rho$):
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2 \right).
\end{equation}

This is described in some detail
by Kendon et al \cite{viv}. The first two terms represent the bulk
contribution, whereas the term in $\kappa$ penalisies curvature in
the interface.

@@

The parameter $\kappa$, which is positive.

\subsubsection{Brazovskii free energy}

This is similar to the symetric, but with one extra term which
requires description.
\begin{equation}
 F[\phi] = 
\int dr \left(
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
+ {\textstyle \frac{1}{2}} C (\nabla^2 \phi)^2 \right).
\end{equation}

\subsubsection{Surfactant model free energy}

\inputkey{free\_energy surfactant}

This is for use with a two order parameter model the first of
which is the composition, as for the symmetric free energy,
while the second $\psi$ represents surfactant concentration.
The free energy density is made up of a number of terms:

\begin{equation}
{\textstyle \frac{1}{2}}A\phi^2
+ {\textstyle \frac{1}{4}}B\phi^4
+ {\textstyle \frac{1}{2}}\kappa (\mathbf{\nabla}\phi)^2
\end{equation}
is the standard symmetric term related to the composition as above;

\begin{equation}
D \left(\psi \ln\psi + (1 - \psi) \ln(1-\psi)\right)
\end{equation}
represents the energy of surfactant in the bluk phase with surfactant
concentration $0 < \psi < 1$ with a single parameter $D$;
\begin{equation}
-{\textstyle \frac{1}{2}} \epsilon \psi (\mathbf{\nabla} \phi)^2
-{\textstyle \frac{1}{2}} \beta \psi^2  (\mathbf{\nabla} \phi)^2
\end{equation}
where the term in $\epsilon$ represents the energy reduction by
adsorbing surfactant at the interface and the term in $\beta$
provides [TO BE ACCOUNTED FOR]; finally, there is an extra term
\begin{equation}
+{\textstyle \frac{1}{2}} W \psi \phi^2
\end{equation}
which is included to stabilise the interface \cite{theissen}.

The corresponding keys for the input file are:

\inputkey{D}

is the bulk surfactant contribution parameter

\inputkey{epsilon}

is the parameter $\epsilon$

\inputkey{W}

is the parameter $W$

\inputkey{beta}

is the value of the parameter $\beta$ (not implemented yet).

\inputkey{phi\_b}

the initial, uniform, background surfactant concentration $\psi_b$.

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
\hline
Name & $\eta$ & $-A,B$ & $\kappa$ & $D$ & $\epsilon$ & $W$
     & $M$ & $M_\psi$ \\
\hline
Run001 & 0.025 & 0.000625 & 0.004 & 0.0001 & 0.004 & 0.0003  & 2.0 & 0.1\\
\hline
\end{tabular}
\caption{Acceptable parameters for the surfactant free energy model.
Note that $\epsilon = \kappa$.}
\end{center}
\end{table}

\subsection{Other Binary Fluid Parameters}

@@

Sets the compositional order parameter mobility $M$ (uniform in space)
which appears in the Cahn-Hilliard equation.

\inputkey{mobilty\_psi}

Sets the (uniform) mobility for the surfactant as appropriate for
the Cahn-Hilliard equation.

@@

The mean compositional order parameter roughly $-0.5 < \phi_0 < 0.5$.
The default value is zero, i.e., a symmetric 50:50 mixture by volume.

\inputkey{phi\_initialisation}

Determines how the compositional order parameter is initialisated
at the start of the run. If set to \texttt{spinodal} the value
is set to $\phi_0$ as set above plus or minus a random noise, the
magnitude of which is set by the value of the \texttt{noise} key.
If set to \texttt{block} a one-dimensional profile is set up in
the $z$-direction representing two blocks of fluid with
$\phi = \pm 1$. The two interfaces are set at $z = L_z/4$ and
$z = 3L_z/4$ with the equilibrium $\tanh(z/\xi_0)$ profile having 
appropriate width. The $\phi = -1$ section is in the middle.

@@

An analtyical expression is also available for the conductance
of a square or rectugular capillary. For an infinite capillary
of square cross section width $w \times h$ (where $h$ is the
longer),
the volume flux per unit area $J$ is expected to be
\cite{papanastasiou,edo1}
\begin{equation}
J = - \frac{1}{3\eta} \frac{\partial p}{\partial x} (h/2)^2
\left[
1 - 6(h/w) \sum_{k=1}^{\infty} \frac{\tanh(\alpha_k w/h)}{\alpha_k^5} 
\right]
\end{equation}
where $\alpha_k = (2k - 1)\pi/2$. The pressure gradient
$-\partial p / \partial x$ is replaced by the body force $\rho g$ and
one can difine the viscosity-independnet conductance $C$ via
$J = C\rho g / \eta$. The calculation procedes as above.
